---
title: "Reviewing the half-clean data with a Hardy-Weinberg and Structure Analysis"
author: "Grace Saville"
date: "16/05/2022"
output: pdf_document
---

```{r setup, include = FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, warning = FALSE)
opts_knit$set(root.dir = "C:/Users/airhe/OneDrive/Documents/Masters/Project 3/kiore-project")
```

```{r libraries, include = FALSE}
library(reshape)
library(plyr)
# getwd()
# setwd("C:/Users/airhe/OneDrive/Documents/Masters/Project 3/kiore-project")
```

# 1. Prepping df for HWE Analysis

I would like to remove the SNPs not in Hardy-Weinberg equilibrium, therefore I need to reformat the data for input into HWE program.

```{r loading data}
# load(".RData") # if necessary
data <- read.csv("./data/RStudio/ratsSNPs_halfclean.csv")
```

## 1a. Reformatting the dataframe
```{r SNP-column-separation}
copy <- data # making a copy
t(copy[1,1:20]) # checking column names
copy <- copy[,-c(2:16)] # removing all but specimen names and SNPs
# copy[1,1:20] # checking

copy[copy == "?"] <- "?:?" # replacing single ? with double ? so alleles can be split

x <- data.frame(island = copy$island) # setting up new df for for loop
coln <- as.vector(colnames(copy)) # prepping to paste the column names into the for loop
dim(copy) # 379 rows 283 columns
for (i in 2:283) {
  y <- colsplit(copy[,i], split = ":", names = c(coln[i], paste("blank", i, sep = "."))) # splitting each i column and renaming them 
  x <- cbind(x, y) # combining output with current df
  rm(i, y) # removing temp objects
}

# Checking:
# dim(x3) # 379 rows 565 columns
# x2[1:5,1:5] 
# x3[1:5,1:5] # comparing the 2 dfs to check the column naming worked correctly

copy <- x
rm(x, coln) # removing excess objects
```

## 1b. Producing the file necessary for PGDSpider program

```{r file-prep}
copy <- copy[order(copy$island, decreasing = FALSE), ] # ordering df alphabetically by island
# as.matrix(copy[, 1]) # printing the island names and row numbers

# A=1, T=2, G=3, C=4
copy[copy == "A"] <- "1"
copy[copy == "T"] <- "2"
copy[copy == "G"] <- "3"
copy[copy == "C"] <- "4"

# row numbers in dataset df listed below for each popn.
popnames <- as.character(
  c(
    "pop = Aotea", # 1:10
    "pop = Borneo", # 11:28
    "pop = Doubtful_Sound", # 315
    "pop = Great_Mercury_Island", # 30
    "pop = Halmahera", # 31:42
    "pop = Hatutaa", # 43:63
    "pop = Honuea", # 64:83
    "pop = Kaikura_Island", # 84:103
    "pop = Kamaka", # 104:124
    "pop = Kayangel", # 125:145
    "pop = Late_Island", # 148:168
    "pop = Mainland", # 29, 146, 147, 169, 358, 359 (including Luzon here because 
    # Luzon is part of the mainland cluster in the NeighborNet network)
    "pop = Malenge", # 170:181
    "pop = Mohotani", # 182:195
    "pop = Motukawanui", # 196:216
    "pop = New_Britain", # 217:226
    "pop = New_Guinea", # 227:229
    "pop = Normanby_Island", # 230
    "pop = Rakiura", # 231:251
    "pop = Reiono", # 252:272
    "pop = Rimatuu", # 273:293
    "pop = Slipper_Island", # 294:314
    "pop = Sulawesi", # 316:337
    "pop = Tahanea", # 338:357
    "pop = Wake_Island" # 360:379
  )
)

# Creating population dfs
a <- as.data.frame(copy[1:10,]) # Aotea
b <- as.data.frame(copy[11:28,]) # Borneo
c <- as.data.frame(copy[315,]) # Doubtful_Sound
d <- as.data.frame(copy[30,]) # Great_Mercury_Island
e <- as.data.frame(copy[31:42,]) # Halmahera
f <- as.data.frame(copy[43:63,]) # Hatutaa
g <- as.data.frame(copy[64:83,]) # Honuea
h <- as.data.frame(copy[84:103,]) # Kaikura_Island
i <- as.data.frame(copy[104:124,]) # Kamaka
j <- as.data.frame(copy[125:145,])  # Kayangel
k <- as.data.frame(copy[148:168,]) # Late_Island
l <- as.data.frame(copy[c(29, 146, 147, 169, 358, 359),]) # Mainland
m <- as.data.frame(copy[170:181,]) # Malenge
n <- as.data.frame(copy[182:195,]) # Mohotani
o <- as.data.frame(copy[196:216,]) #  Motukawanui
p <- as.data.frame(copy[217:226,]) #  New_Britain
q <- as.data.frame(copy[227:229,]) # New_Guinea
r <- as.data.frame(copy[230,]) #  Normanby_Island
s <- as.data.frame(copy[231:251,]) # Rakiura
t <- as.data.frame(copy[252:272,]) # Reiono
u <- as.data.frame(copy[273:293,]) # Rimatuu
v <- as.data.frame(copy[294:314,]) # Slipper_Island
w <- as.data.frame(copy[316:337,]) #  Sulawesi
x <- as.data.frame(copy[338:357,]) # Tahanea
y <- as.data.frame(copy[360:379,]) # Wake_Island

pops <- as.character(c(letters[seq(from = 1, to = 25)])) # list of popn object names
```

```{r writing-to-file, eval = FALSE}
ncol(copy) #565
getwd()

sink("./data/ratsSNPs_PGDSpider_input.txt") # create empty file
cat("rats_SNPS", "npops = 25", "nloci = 282", fill = 1)
cat("\t", fill = FALSE)
cat(colnames(copy[,c(FALSE,TRUE)]), "\n", sep = "\t\t", fill = FALSE) # column/SNP 
# names (even columns only)
for (i1 in 1:25) {
  cat(popnames[i1], fill = 1) # island name
  foo <- get(pops[i1]) # calling the island object based on the pops vector
  for (i2 in 1:nrow(foo)) {
    cat(as.character(foo[i2, ]), "\n", fill = FALSE, sep = "\t") # print SNP rows
  } # inner loop close
} # outer loop close
sink() # closing the sink connection (do not forget!)
```

```{r rm}
rm(i1, i2, foo, popnames, pops)
rm(list = c(letters[seq(from = 1, to = 25)])) # removing excess objects
```

At this stage PGDSpider program and Arlequin were used to convert the file produced and run tests on the data. The resulting output is used here for analysis. Full method detailled in the README file.

# 2. HWE Analysis and removal

## 2a. Creating loop for reading the HWE files
```{r reading-HWE-files}
filenames <- as.vector(list.files("./results/Arlequin_HardyWeinberg/hwe_results_by_island_14032022"))

# START OF MEGA FOR LOOP:
for (i in 1:length(filenames)) {
  df <- read.delim(paste0("./results/Arlequin_HardyWeinberg/hwe_results_by_island_14032022/", filenames[i]))
  m <- as.vector(grep("This locus is monomorphic", df[,1], value = FALSE, fixed = TRUE)) 
  # making list of rows that only say the above words
  df <- as.data.frame(df[-c(1,m),]) # removing the rows listed above, plus the dashed line
  
  df <- as.data.frame(gsub("  ", " ", df[,1], fixed = TRUE)) # removing spaces
  df <- as.data.frame(gsub("  ", " ", df[,1], fixed = TRUE)) # removing spaces
  df <- as.data.frame(gsub("  ", " ", df[,1], fixed = TRUE)) # removing spaces
  
  colnames(df) <- "Var1"
  df <- tidyr::separate(df, sep = " ", col = Var1, into = c("foo", "Locus", "Genot", "Obs.Het", "Exp.Het", "P.value", "s.d", "Steps.done")) # splitting column into multiple (with a extra column because of the space at the start of the string)
  df <- df[,-1] # removing extra row
  for (ii in 1:ncol(df)) {df[,ii] <- as.numeric(df[,ii])} # converting to numeric
  assign(paste(filenames[i]), df) # renaming object
  # write.table(df, paste("df", filenames[i], sep = "_"), row.names = FALSE, sep = "\t") # save to file if necessary
} # END OF FOR LOOP

rm(i, ii, m, df, filenames)
# setwd("C:/Users/airhe/OneDrive/Documents/Masters/Project 3/kiore-project")
```

## 2b. Checking the HWE P-values

```{r making-hwe-results-df}
objectnames <- as.vector(ls()) # should be islands only, otherwise remove extras from vector
objectnames
# If necessary:
objectnames <- objectnames[-c(3, 4)] # removing non-island objects, may not be the same numbers

# making df of all hwe results
hwe.all <- data.frame()
for (i in 1:length(objectnames)) {
  foo <- get(objectnames[i])
  islandpop <- c(rep(paste(objectnames[i]), paste(nrow(foo)))) # making a vector of the popn. name
  foo$islandpop <- islandpop # adding the column to the results df to identify popn.
  hwe.all <- rbind(hwe.all, foo) # adding the popn. df to the combined hwe results df
}

rm(islandpop, foo, i)
rm(list = objectnames) # removes all the island objects
```

## 2c. Running Holm's Sequential Bonferroni test to adjust p-values

```{r bonferroni-test}
nrow(hwe.all) # 2622

p.value.adjusted <- c(p.adjust(hwe.all$P.value, method = "holm")) # adjusting p-values
hwe.all$p.value.adjusted <- p.value.adjusted # making new column

rm(p.value.adjusted)
# getwd()
# write.csv(hwe.all, "./results/Arlequin_HardyWeinberg/HWEanalysis_allresults.csv", row.names = FALSE)
```

## 2d. Examining significant hwe p-values

```{r adjusted-p-value-summary}
hwe.all <- read.csv("./results/Arlequin_HardyWeinberg/HWEanalysis_allresults.csv")
hwe.signif <- hwe.all[which(hwe.all$p.value.adjusted <= 0.05),]
kable(hwe.signif[,c(1,8,9)])
kable(count(hwe.signif$Locus)) # 2 at locus 41 (Kaikura and Reiono), rest are singles
kable(count(hwe.signif$islandpop)) # concerning that 17 of 23 are Kayangel
```

# 3. Removing samples/loci with issues identified in HWE and Structure analyses
```{r specimen-removal}
# getwd()
halfclean <- read.csv("./data/RStudio/ratsSNPs_halfclean.csv")

# need to remove Kamaka_009
# not removing Rimatuu_19 and Rimatuu_20 because they are right next to Reiono
# and could have swam.
# both the (pre-cleanup) NeighborNet and Structure identify Kayangel17 as concerning, 
# as well as Kayangel11, 13, 15, 19, and 21. HWE shows also several loci in Kayangel 
# as problematic, but not in other popn.s (with 1 exception). Since the loci are 
# only problematic in Kayangel popn., I believe the issue is in the specimens, not 
# the loci themselves.

remove <- c(
    "Kamaka_009",
    "Rimatuu_19", # actually should re-do analysis keeping the rimatuu's!
    "Rimatuu_20",
    "Kayangel11",
    "Kayangel13",
    "Kayangel15",
    "Kayangel17",
    "Kayangel19",
    "Kayangel21"
  ) # names of specimens to remove (each name should have exactly 10 characters)
x <- sapply(remove, function(i) grep(i, x = halfclean$island, value = FALSE)) 
# finding the row numbers of the specimens to remove

halfclean[c(x),1] # checking the names match
clean <- halfclean[-c(x),] # removing rows described above

# getwd()
# write.csv(clean, "./data/RStudio/ratsSNPs_clean.csv", row.names = FALSE)

rm(x, remove)
```

# 4. Double checking for monomorphic columns (SNP loci)

```{r invariant-column-check}
ncol(clean) #298
monocols <- integer() # empty vector for the for loop
for (i in 17:298) {
  z <- length(unique(clean[,i])) # no. of unique values in the row (looking for 1, or 2 if there's "?")
    if (z <= 3) 
      {monocols <- append(monocols, i) # if z is as so, add the column number to the vector
    }
  rm(z)
} 
# tried with z <= 2 but no result, therefore tried z <= 3 and checked the results manually below.

monocols
for (i in monocols) {
  print(unique(clean[,i]))
}
# none with only 1 unique SNP in each column. It's possible since the SNP loci were selected for their differences, but double check this 
```

```{r possible-column-deletion-2}
# x <- x[,-c(monocols)] # for removal of monomorphic columns, but none found

rm(i, monocols)
```

# 5. Summary: specimens per Island after full data clean-up

```{r count-per-island-2}
kable(count(clean$island.1))
```

|Island|freq before cleanup|freq after cleanup|difference|
|:----:|:----:|:----:|:----:|
|Aotea (Great Barrier I)|10|10|0|
|Borneo|25|18|7|
|Doubtful Sound|1|1|0|
|Great Mercury Island|1|1|0|
|Halmahera|25|12|13|
|Hatutaa|21|21|0|
|Honuea|21|20|1|
|Kaikura Island|20|20|0|
|Kamaka|21|20|1|
|Kayangel|21|15|6|
|Late Island|21|21|0|
|Luzon|1|1|0|
|Mainland|5|5|0|
|Malenge|25|12|13|
|Mohotani|14|14|0|
|Motukawanui|21|21|0|
|New Britain|26|10|16|
|New Guinea|25|3|22|
|Normanby Island|25|1|24|
|Rakiura (Stewart Isl)|21|21|0|
|Reiono|21|21|0|
|Rimatuu (Tetiaroa)|21|21|0|
|Slipper Island|21|21|0|
|Sulawesi|25|22|3|
|Tahanea|20|20|0|
|Wake Island|20|20|0|

Islands represented by very few specimens (<= 3) are Doubtful Sound, Great Mercury Island, Luzon, New Guinea, and Normanby Island. 