---
title: "Reviewing the half-clean data with a Hardy-Weinberg and Structure Analysis"
author: "Grace Saville"
date: "19/04/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

# Preamble 

```{r libraries}
library(reshape)
library()
getwd()
setwd("C:/Users/airhe/OneDrive/Documents/Masters/Project 3/kiore-project")
```

# Prepping df for HWE Analysis

I would like to remove the SNPs not in Hardy-Weinberg equilibrium, therefore I need to reformat the data for input into HWE program.

```{r loading data}
# load(".RData") # if necessary
data <- read.csv("./data/RStudio/ratsSNPs_halfclean.csv")
```

```{r SNP column separation}
copy <- data # making a copy
copy[1,1:20] # checking column names
copy <- copy[,-c(2:16)] # removing all but specimen names and SNPs
copy[1,1:20] # checking

copy[copy == "?"] <- "?:?" # replacing single ? with double ? so alleles can be split

x <- data.frame(island = copy$island) # setting up new df for for loop
coln <- as.vector(colnames(copy)) # prepping to paste the column names into the for loop
dim(copy) # 379 rows 283 columns
for (i in 2:283) {
  y <- colsplit(copy[,i], split = ":", names = c(coln[i], paste("blank", i, sep = "."))) # splitting each i column and renaming them 
  x <- cbind(x, y) # combining output with current df
  rm(i, y) # removing temp objects
}

# Checking:
# dim(x3) # 379 rows 565 columns
# x2[1:5,1:5] 
# x3[1:5,1:5] # comparing the 2 dfs to check the column naming worked correctly

copy <- x
rm(x, coln) # removing excess objects
```

### Producing the file necessary for PGDSpider program

```{r file prep}
copy <- copy[order(copy$island, decreasing = FALSE), ] # ordering df alphabetically by island
print(as.matrix(copy[, 1])) # printing the island names and row numbers

# A=1, T=2, G=3, C=4
copy[copy == "A"] <- "1"
copy[copy == "T"] <- "2"
copy[copy == "G"] <- "3"
copy[copy == "C"] <- "4"

# row numbers in dataset df listed below for each popn.
popnames <- as.character(
  c(
    "pop = Aotea", # 1:10
    "pop = Borneo", # 11:28
    "pop = Doubtful_Sound", # 315
    "pop = Great_Mercury_Island", # 30
    "pop = Halmahera", # 31:42
    "pop = Hatutaa", # 43:63
    "pop = Honuea", # 64:83
    "pop = Kaikura_Island", # 84:103
    "pop = Kamaka", # 104:124
    "pop = Kayangel", # 125:145
    "pop = Late_Island", # 148:168
    "pop = Mainland", # 29, 146, 147, 169, 358, 359 (including Luzon here)
    "pop = Malenge", # 170:181
    "pop = Mohotani", # 182:195
    "pop = Motukawanui", # 196:216
    "pop = New_Britain", # 217:226
    "pop = New_Guinea", # 227:229
    "pop = Normanby_Island", # 230
    "pop = Rakiura", # 231:251
    "pop = Reiono", # 252:272
    "pop = Rimatuu", # 273:293
    "pop = Slipper_Island", # 294:314
    "pop = Sulawesi", # 316:337
    "pop = Tahanea", # 338:357
    "pop = Wake_Island" # 360:379
  )
)

# Creating population dfs
a <- as.data.frame(copy[1:10,]) # Aotea
b <- as.data.frame(copy[11:28,]) # Borneo
c <- as.data.frame(copy[315,]) # Doubtful_Sound
d <- as.data.frame(copy[30,]) # Great_Mercury_Island
e <- as.data.frame(copy[31:42,]) # Halmahera
f <- as.data.frame(copy[43:63,]) # Hatutaa
g <- as.data.frame(copy[64:83,]) # Honuea
h <- as.data.frame(copy[84:103,]) # Kaikura_Island
i <- as.data.frame(copy[104:124,]) # Kamaka
j <- as.data.frame(copy[125:145,])  # Kayangel
k <- as.data.frame(copy[148:168,]) # Late_Island
l <- as.data.frame(copy[c(29, 146, 147, 169, 358, 359),]) # Mainland
m <- as.data.frame(copy[170:181,]) # Malenge
n <- as.data.frame(copy[182:195,]) # Mohotani
o <- as.data.frame(copy[196:216,]) #  Motukawanui
p <- as.data.frame(copy[217:226,]) #  New_Britain
q <- as.data.frame(copy[227:229,]) # New_Guinea
r <- as.data.frame(copy[230,]) #  Normanby_Island
s <- as.data.frame(copy[231:251,]) # Rakiura
t <- as.data.frame(copy[252:272,]) # Reiono
u <- as.data.frame(copy[273:293,]) # Rimatuu
v <- as.data.frame(copy[294:314,]) # Slipper_Island
w <- as.data.frame(copy[316:337,]) #  Sulawesi
x <- as.data.frame(copy[338:357,]) # Tahanea
y <- as.data.frame(copy[360:379,]) # Wake_Island

pops <- as.character(c(letters[seq(from = 1, to = 25)])) # list of popn object names
```

```{r writing to file}
ncol(copy) #565
getwd()

sink("./data/ratsSNPs_PGDSpider_input.txt") # create empty file
cat("rats_SNPS", "npops = 25", "nloci = 282", fill = 1)
cat("\t", fill = FALSE)
cat(colnames(copy[,c(FALSE,TRUE)]), "\n", sep = "\t\t", fill = FALSE) # column/SNP names (even columns only)
for (i1 in 1:25) {
  cat(popnames[i1], fill = 1) # island name
  foo <- get(pops[i1]) # calling the island object based on the pops vector
  for (i2 in 1:nrow(foo)) {
    cat(as.character(foo[i2, ]), "\n", fill = FALSE, sep = "\t") # printing the SNP rows
  } # inner loop close
} # outer loop close
sink() # closing the sink connection (do not forget!)

rm(i1, i2, foo, popnames, pops)
rm(list = c(letters[seq(from = 1, to = 25)])) # removing excess objects
```

# HWE Analysis and removal

### Creating loop for reading the HWE files
```{r reading HWE files}
getwd()
setwd("./results/Arlequin_HardyWeinberg/hwe_results_by_island_14032022")
filenames <- as.vector(list.files())

for (i in 1:length(filenames)) {
  df <- read.delim(filenames[i])
  m <- as.vector(grep("This locus is monomorphic", df[,1], value = FALSE, fixed = TRUE)) # making list of rows that only say these words
  df <- as.data.frame(df[-c(1,m),]) # removing the rows listed above, plus the dashed line
  
  df <- as.data.frame(gsub("  ", " ", df[,1], fixed = TRUE)) # removing spaces
  df <- as.data.frame(gsub("  ", " ", df[,1], fixed = TRUE)) # removing spaces
  df <- as.data.frame(gsub("  ", " ", df[,1], fixed = TRUE)) # removing spaces
  
  colnames(df) <- "Var1"
  df <- tidyr::separate(df, sep = " ", col = Var1, into = c("foo", "Locus", "Genot", "Obs.Het", "Exp.Het", "P.value", "s.d", "Steps.done")) # splitting column into multiple (with a extra column because of the space at the start of the string)
  df <- df[,-1] # removing extra row
  for (ii in 1:ncol(df)) {df[,ii] <- as.numeric(df[,ii])} # converting to numeric rather than character/factor
  assign(paste(filenames[i]), df) # renaming object
  # write.table(df, paste("df", filenames[i], sep = "_"), row.names = FALSE, sep = "\t") # save to file
}

rm(i, ii, m, df, filenames)
setwd("C:/Users/airhe/OneDrive/Documents/Masters/Project 3/kiore-project")
```

### Checking the HWE P-values

```{r making hwe results df}
objectnames <- as.vector(ls()) # should be islands only, otherwise remove extras from vector
objectnames
# If necessary:
objectnames <- objectnames[-c(3, 27, 28)] # removing non-island objects

# making df of all hwe results
hwe.all <- data.frame()
for (i in 1:length(objectnames)) {
  foo <- get(objectnames[i])
  islandpop <- c(rep(paste(objectnames[i]), paste(nrow(foo)))) # making a vector of the popn. name
  foo$islandpop <- islandpop # adding the column to the results df to identify popn.
  hwe.all <- rbind(hwe.all, foo) # adding the popn. df to the combined hwe results df
}

rm(islandpop, foo, i)
rm(list = objectnames) # removes all the island objects
```

### Running Holm's Sequential Bonferroni test to adjust p-values
```{r bonferroni test}
nrow(hwe.all) # 2622

p.value.adjusted <- c(p.adjust(hwe.all$P.value, method = "holm")) # adjusting p-values
hwe.all$p.value.adjusted <- p.value.adjusted # making new column

rm(p.value.adjusted)
getwd()
write.csv(hwe.all, "./results/Arlequin_HardyWeinberg/HWEanalysis_allresults.csv", row.names = FALSE)
```

### Examining significant hwe p-values
```{r adjusted p-value summary}
hwe.all <- read.csv("./results/Arlequin_HardyWeinberg/HWEanalysis_allresults.csv")
hwe.signif <- hwe.all[which(hwe.all$p.value.adjusted <= 0.05),]
hwe.signif[,c(1,8,9)]
plyr::count(hwe.signif$Locus) # 2 at locus 41 (Kaikura and Reiono), rest are singles
plyr::count(hwe.signif$islandpop) # concerning that 17 of 23 are Kayangel
```

### Adjusted p-values per Locus

|Locus|islandpop|p.value.adjusted|
|:----:|:----:|:----:|
|127|honuea|0.02613|
|182|honuea|0.00000|
|41|kaikura_island|0.02613|
|16|kayangel|0.02613|
|37|kayangel|0.02613|
|50|kayangel|0.02613|
|54|kayangel|0.02613|
|61|kayangel|0.02613|
|62|kayangel|0.00000|
|67|kayangel|0.00000|
|84|kayangel|0.00000|
|93|kayangel|0.02613|
|122|kayangel|0.00000|
|170|kayangel|0.02613|
|171|kayangel|0.02613|
|177|kayangel|0.00000|
|211|kayangel|0.02613|
|252|kayangel|0.02613|
|266|kayangel|0.00000|
|278|kayangel|0.00000|
|107|rakiura|0.02613|
|128|rakiura|0.02613|
|41|reiono|0.00000|

### Number of islands with a significant adjusted p-value at a particular locus

|loci column|freq|
|:----:|:----:|
|16|1|
|37|1|
|41|2|
|50|1|
|54|1|
|61|1|
|62|1|
|67|1|
|84|1|
|93|1|
|107|1|
|122|1|
|127|1|
|128|1|
|170|1|
|171|1|
|177|1|
|182|1|
|211|1|
|252|1|
|266|1|
|278|1|

### Number of loci per island population with significant adjusted p-values

|island population|no. of signif. loci|
|:----:|:----:|
|honuea|2|
|kaikura_island|1|
|kayangel|17|
|rakiura|2|
|reiono|1|

# Removing samples/loci with issues identified in HWE and Structure analyses
```{r specimen removal}
getwd()
halfclean <- read.csv("./data/RStudio/ratsSNPs_halfclean.csv")

# need to remove Kamaka_009, and Rimatuu_19 and Rimatuu_20 due to position in Structure.
# both the (pre-cleanup) NeighborNet and Structure identify Kayangel17 as concerning, as well as Kayangel11, 13, 15, 19, and 21. HWE shows also several loci in Kayangel as problematic, but not in other popn.s (with 1 exception). Since the loci are only problematic in Kayangel popn., I believe the issue is in the specimens, not the loci themselves.

remove <- c(
    "Kamaka_009",
    "Rimatuu_19",
    "Rimatuu_20",
    "Kayangel11",
    "Kayangel13",
    "Kayangel15",
    "Kayangel17",
    "Kayangel19",
    "Kayangel21"
  ) # names of specimens to remove (each name should have exactly 10 characters)
x <- sapply(remove, function(i) grep(i, x = halfclean$island, value = FALSE)) # finding the row numbers of the specimens to remove
halfclean[c(x),1] # checking the names match
clean <- halfclean[-c(x),] # removing rows described above

getwd()
write.csv(clean, "./data/RStudio/ratsSNPs_clean.csv", row.names = FALSE)

rm(x, remove)
```

# Double checking for monomorphic columns (SNP loci)

```{r invariant column check}
ncol(clean) #298
monocols <- integer() # empty vector for the for loop
for (i in 17:298) {
  z <- length(unique(clean[,i])) # no. of unique values in the row (looking for 1, or 2 if there's "?")
    if (z <= 3) 
      {monocols <- append(monocols, i) # if z is as so, add the column number to the vector
    }
  rm(z)
} 
# tried with z <= 2 but no result, therefore tried z <= 3 and checked the results manually below.

monocols # 29  30  43  52  54  76  84  86  89 105 113 115 123 127 136 149 154 155 156 170 179 182 183 200 204 205 209 211 214 216 222 223 224 230 235 243 290 295 296
for (i in monocols) {
  print(count(clean[,i]))
}
# none with only 1 unique SNP in each column. It's possible since the SNP loci were selected for their differences, but double check this 
```

```{r possible column deletion 2}
# x <- x[,-c(monocols)] # for removal of monomorphic columns, but none found

rm(i, monocols)
```

# Specimens per Island after full data clean-up

```{r count per island 2}
plyr::count(clean$island.1)
```

|Island|freq before cleanup|freq after cleanup|difference|
|:----:|:----:|:----:|:----:|
|Aotea (Great Barrier I)|10|10|0|
|Borneo|25|18|7|
|Doubtful Sound|1|1|0|
|Great Mercury Island|1|1|0|
|Halmahera|25|12|13|
|Hatutaa|21|21|0|
|Honuea|21|20|1|
|Kaikura Island|20|20|0|
|Kamaka|21|20|1|
|Kayangel|21|15|6|
|Late Island|21|21|0|
|Luzon|1|1|0|
|Mainland|5|5|0|
|Malenge|25|12|13|
|Mohotani|14|14|0|
|Motukawanui|21|21|0|
|New Britain|26|10|16|
|New Guinea|25|3|22|
|Normanby Island|25|1|24|
|Rakiura (Stewart Isl)|21|21|0|
|Reiono|21|21|0|
|Rimatuu (Tetiaroa)|21|19|2|
|Slipper Island|21|21|0|
|Sulawesi|25|22|3|
|Tahanea|20|20|0|
|Wake Island|20|20|0|

Islands represented by very few specimens (<= 3) are Doubtful Sound, Great Mercury Island, Luzon, New Guinea, and Normanby Island. 