---
title: "FST Prep and Analysis Script"
author: "Grace Saville"
date: "16/05/2022"
output: pdf_document
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE)
opts_knit$set(root.dir = "C:/Users/airhe/OneDrive/Documents/Masters/Project 3/kiore-project")
```

```{r libraries, include = FALSE}
library(corrplot)
library(ggplot2)
library(tidyr)
library(car)
library(geosphere)
library(ade4)
# getwd()
# setwd("C:/Users/airhe/OneDrive/Documents/Masters/Project 3/kiore-project")
```

# 1. Prepping the dataframe

```{r loading-data}
# load(".RData") # if necessary
data <- read.csv("./data/RStudio/ratsSNPs_clean.csv")
```

```{r SNP-column-separation, warning = FALSE}
copy <- data # making a copy
t(copy[1,1:20]) # checking column names
copy <- copy[,-c(2:16)] # removing all but specimen names and SNPs
t(copy[1,1:20]) # checking

copy[copy == "?"] <- "?:?" # replacing single ? with double ? so alleles can be split

x <- data.frame(island = copy$island) # setting up new df for for loop
coln <- as.vector(colnames(copy)) # prepping to paste the column names into the for loop
dim(copy) # 379 rows 283 columns
for (i in 2:283) {
  y <- reshape::colsplit(
          copy[, i], split = ":", names = c(coln[i], paste("blank", i, sep = "."))) 
          # splitting each i column and renaming them
  x <- cbind(x, y) # combining output with current df
  rm(i, y) # removing temp objects
}

# Checking:
# dim(x3) # 379 rows 565 columns
# x2[1:5,1:5] 
# x3[1:5,1:5] # comparing the 2 dfs to check the column naming worked correctly

copy <- x
rm(x, coln) # removing excess objects
```

# 2. Producing the file necessary for PGDSpider program

```{r file-prep}
copy <- copy[order(copy$island, decreasing = FALSE), ] # ordering df alphabetically 
# by island
# print(as.matrix(copy[, 1])) # printing the island names and row numbers

# A=1, T=2, G=3, C=4
copy[copy == "A"] <- "1"
copy[copy == "T"] <- "2"
copy[copy == "G"] <- "3"
copy[copy == "C"] <- "4"

# row numbers in dataset df listed below for each popn.
popnames <- as.character(
  c(
    "pop = Aotea", # 1:10
    "pop = Borneo", # 11:28
    "pop = Doubtful_Sound", # 306
    "pop = Great_Mercury_Island", # 30
    "pop = Halmahera", # 31:42
    "pop = Hatutaa", # 43:63
    "pop = Honuea", # 64:83
    "pop = Kaikura_Island", # 84:103
    "pop = Kamaka", # 104:123
    "pop = Kayangel", # 124:138
    "pop = Late_Island", # 141:161
    "pop = Mainland", # 29, 139, 140, 162, 349, 350 (including Luzon here because
    # Luzon is part of the mainland cluster in the NeighborNet network)
    "pop = Malenge", # 163:174
    "pop = Mohotani", # 175:188
    "pop = Motukawanui", # 189:209
    "pop = New_Britain", # 210:219
    "pop = New_Guinea", # 220:221
    "pop = Normanby_Island", # 223
    "pop = Rakiura", # 224:244
    "pop = Reiono", # 245:265
    "pop = Rimatuu", # 266:284
    "pop = Slipper_Island", # 285:305
    "pop = Sulawesi", # 307:328
    "pop = Tahanea", # 329:348
    "pop = Wake_Island" # 351:370
  )
)

# Creating population dfs
a <- as.data.frame(copy[1:10,]) # Aotea
b <- as.data.frame(copy[11:28,]) # Borneo
c <- as.data.frame(copy[306,]) # Doubtful_Sound
d <- as.data.frame(copy[30,]) # Great_Mercury_Island
e <- as.data.frame(copy[31:42,]) # Halmahera
f <- as.data.frame(copy[43:63,]) # Hatutaa
g <- as.data.frame(copy[64:83,]) # Honuea
h <- as.data.frame(copy[84:103,]) # Kaikura_Island
i <- as.data.frame(copy[104:123,]) # Kamaka
j <- as.data.frame(copy[124:138,])  # Kayangel
k <- as.data.frame(copy[141:161,]) # Late_Island
l <- as.data.frame(copy[c(29, 139, 140, 162, 349, 350),]) # Mainland
m <- as.data.frame(copy[163:174,]) # Malenge
n <- as.data.frame(copy[175:188,]) # Mohotani
o <- as.data.frame(copy[189:209,]) #  Motukawanui
p <- as.data.frame(copy[210:219,]) #  New_Britain
q <- as.data.frame(copy[220:221,]) # New_Guinea
r <- as.data.frame(copy[223,]) #  Normanby_Island
s <- as.data.frame(copy[224:244,]) # Rakiura
t <- as.data.frame(copy[245:265,]) # Reiono
u <- as.data.frame(copy[266:284,]) # Rimatuu
v <- as.data.frame(copy[285:305,]) # Slipper_Island
w <- as.data.frame(copy[307:328,]) #  Sulawesi
x <- as.data.frame(copy[329:348,]) # Tahanea
y <- as.data.frame(copy[351:370,]) # Wake_Island

pops <- as.character(c(letters[seq(from = 1, to = 25)])) # list of popn object names
```

```{r writing-file, eval = FALSE}
ncol(copy) #565
getwd()

sink("./data/PGDSpider/ratsSNPs_PGDSpider_input_CLEAN.txt") # create empty file
cat("rats_SNPS", "npops = 25", "nloci = 282", fill = 1)
cat("\t", fill = FALSE)
cat(colnames(copy[,c(FALSE,TRUE)]), "\n", sep = "\t\t", fill = FALSE) # column/SNP names
# (even columns only)
for (i1 in 1:25) {
  cat(popnames[i1], fill = 1) # island name
  foo <- get(pops[i1]) # calling the island object based on the pops vector
  for (i2 in 1:nrow(foo)) {
    cat(as.character(foo[i2, ]), "\n", fill = FALSE, sep = "\t") # printing the SNP rows
  } # inner loop close
} # outer loop close
sink() # closing the sink connection (do not forget!)

rm(i1, i2, foo, popnames, pops)
rm(list = c(letters[seq(from = 1, to = 25)])) # removing excess objects
```

At this stage PGDSpider program and Arlequin were used to convert the file produced and run tests on the data. The resulting output is used here for analysis.

# 3. Loading the results files for analysis

```{r loading-results-files}
popnames <- as.character(
  c(
    "Aotea",
    "Borneo",
    "Doubtful_Sound",
    "Great_Mercury_Island",
    "Halmahera",
    "Hatutaa",
    "Honuea",
    "Kaikura_Island",
    "Kamaka",
    "Kayangel",
    "Late_Island",
    "Mainland", # (inc Luzon here)
    "Malenge",
    "Mohotani",
    "Motukawanui",
    "New_Britain",
    "New_Guinea",
    "Normanby_Island",
    "Rakiura",
    "Reiono",
    "Rimatuu",
    "Slipper_Island",
    "Sulawesi",
    "Tahanea",
    "Wake_Island" 
  )
)

pwd <- read.csv("./results/Arlequin_FST/fst_pairwisedistances_only.csv", 
                header = TRUE)
pv <- read.csv("./results/Arlequin_FST/fst_pairwisedistances_pvalues_only.csv", 
               header = TRUE)
colnames(pwd) <- popnames
rownames(pwd) <- popnames
pwd <- as.matrix(pwd)
colnames(pv) <- popnames
rownames(pv) <- popnames
pv <- as.matrix(pv)
```

```{r making-full-matrix}
x <- t(pwd) # transposed copy
pwd[upper.tri(pwd, diag = FALSE)] <- x[upper.tri(x, diag = FALSE)] # making full 
# matrix (not just lower tri)
x <- t(pv) # transposed copy for p-values
pv[upper.tri(pv, diag = FALSE)] <- x[upper.tri(x, diag = FALSE)]

rm(x)
```

# 4. Making Geopgraphic distance matrix

```{r making-geo-distance-matrix}
longlat <- data[,c(8,11,12)]
longlat <- longlat[!duplicated(longlat$island.1),] # keeping only 1 coordinate 
# for each island
longlat <- longlat[order(longlat$island.1, decreasing = FALSE),] # sorting alphabetically
row.names(longlat) <- seq(nrow(longlat)) # renaming row numbers to be sequential
kable(longlat) # checking

# editing the names to match those in the pwd df so I can merge them later
longlat[1,1] <- "Aotea"
longlat[3,1] <- "Doubtful_Sound"
longlat[4,1] <- "Great_Mercury_Island"
longlat[8,1] <- "Kaikura_Island"
longlat[11,1] <- "Late_Island"
longlat[17,1] <- "New_Britain"
longlat[18,1] <- "New_Guinea"
longlat[19,1] <- "Normanby_Island"
longlat[20,1] <- "Rakiura"
longlat[22,1] <- "Rimatuu"
longlat[23,1] <- "Slipper_Island"
longlat[26,1] <- "Wake_Island"
longlat <- longlat[-12,] # removing luzon since fst has it with mainland
row.names(longlat) <- seq(nrow(longlat)) # renaming row numbers to be sequential

geo.matrix <- as.matrix(longlat[,c(3,2)]) # distGeo function needs a matrix with 
# 2 columns, col 1 longitude and col 2 latitude

geo.matrix <- distm(geo.matrix, fun = distGeo) # converting to pairwise distance matrix
dim(geo.matrix) # 25 25
```

# 5. FST Mantel test

```{r mantel-test-fst-geo}
dim(geo.matrix) # 25 25
dim(pwd) # 25 25 

geo.dist <- as.dist(geo.matrix, diag = TRUE, upper = TRUE) # converting to dist object
fst.dist <- as.dist(pwd, diag = TRUE, upper = TRUE) 

set.seed(4)
r1 <- mantel.rtest(fst.dist, geo.dist, nrepet = 999)
r1
plot(r1$plot$hist, main = "Mantel test", xlim = c(-0.5, 0.5))
# abline(v = )
```


# 6. Creating results dataframe on which to base analyses

## 6a. Converting FST matrix to dataframe

```{r fst-matrix-to-df}
pwd.df <- pwd
pwd.df[lower.tri(pwd.df, diag = TRUE)] <- NA # keeping only the upper triangle 
# of each matrix

pwd.df <- data.frame(
  col = colnames(pwd.df)[col(pwd.df)],
  row = rownames(pwd.df)[row(pwd.df)],
  fst.dist = c(pwd.df)
) # converting the fst matrix into a df with columns describing which combos 
# result in the distance

pwd.df <- na.omit(pwd.df)

pwd.df <- unite(pwd.df, islands.combo, 1:2, sep = ":", remove = TRUE) # combining 
# the first 2 columns (the names of the matrices columns and rows) to give a label 
# to each pairwise distance
```

## 6b. Combining the FST and Geographic dataframes

```{r adding-geo-distances}
colnames(geo.matrix) <- longlat[,1]
rownames(geo.matrix) <- longlat[,1] # naming the rows and columns

geo.matrix[lower.tri(geo.matrix, diag = TRUE)] <- NA # keeping only the upper 
# triangle of matrix

geo.df <- data.frame(
  col = colnames(geo.matrix)[col(geo.matrix)],
  row = rownames(geo.matrix)[row(geo.matrix)],
  geo.dist = c(geo.matrix)
) # converting the genetic matrix into a df with columns describing which combos 
# result in the distance

geo.df <- na.omit(geo.df) # removing NA's left from lower triangle

geo.df <- unite(geo.df, islands.combo, 1:2, sep = ":", remove = TRUE) # combining 
# the first 2 columns (the names of the matrices columns and rows) to give a label 
# to each pairwise distance

pwd.df <- merge(pwd.df, geo.df, by = "islands.combo", all = FALSE) # merging distance
# between islands with FST df

pwd.df$geo.dist <- pwd.df$geo.dist/1000 # going from metres to km

rm(geo.df)
```

## 6c. Saving outcomes

```{r saving, eval = FALSE}
write.csv(pwd.df, "./Results/Arlequin_FST/FST_RStudio_outcomes_df.csv", row.names = FALSE)
write.csv(pv, "./Results/Arlequin_FST/FST_RStudio_pvalue_matrix.csv", row.names = FALSE)
```

# 7. Linear modelling

```{r loading-data-if-necessary, eval = FALSE}
pwd.df <- read.csv("./Results/Arlequin_FST/FST_RStudio_outcomes_df.csv")
```

## 7a. Test model 

```{r test-model-creation}
testLM <- lm(fst.dist ~ geo.dist, data = pwd.df) # model
summary(testLM) # model results

plot(fst.dist ~ geo.dist, data = pwd.df)
abline(coef = coef(testLM), col = 4, lwd = 2)
```

```{r test-model-diagnostics}
par(mfrow = c(2, 2)) # changing the number of plots visible at once 
plot(testLM) # diagnostic plots

par(mfrow = c(1,1))
hist(testLM$residuals, breaks = 10, xlim = c(-1,1))

qqPlot(testLM$residuals, line = "quartiles") # normal, possible outliers 132, 148
shapiro.test(testLM$residuals) # indicates non-normality of residuals but 
# likely affected by the 2 outliers mentioned above

ncvTest(testLM) # homoscedasticity test: H0 of constant variance is rejected. 
# Supported by downward slope in Scale-location plot (plot(testLM))

influenceIndexPlot(testLM) # outliers
# Cooks distances: none larger than 0.5, 
# Studentised residuals: 132 and 148 less than -3
# Bonferroni p-value: 132 and 148 smaller then 0.05, 
# Hat-values: none influential, higher than 1
outlierTest(testLM)
```

* Diagnostic Plots: indications that the relationship is linear, normal distribution of residuals, down trending scale-location plot and cone-shaped residuals vs. fitted plot therefore non-constant variance, and 132 (New Guinea to Mainland) and 148 (Normanby Island to Mainland) are potential issues.

## 7b. Adjusted Model

```{r adjusted-model-building}
pwd.df2 <- pwd.df[pwd.df$fst.dist >= 0,] # removing 132 (New_Guinea:Mainland) and 
# 148 (Normanby_Island:Mainland) which are both negative FST values

x <- lm(fst.dist ~ geo.dist, data = pwd.df2) # creating model to take weights from
wt <- 1 / lm(abs(x$residuals) ~ x$fitted.values)$fitted.values^2 # weighting 
# residuals by how large they are
rm(x) 

LM <- lm(fst.dist ~ geo.dist, data = pwd.df2, weights = wt) # weighted residual model
summary(LM) # model results

plot(fst.dist ~ geo.dist, data = pwd.df2)
abline(coef = coef(LM), col = 4, lwd = 2)
```

```{r adjusted-model-diagnostics}
par(mfrow = c(2, 2)) # changing the number of plots visible at once
plot(LM) # diagnostic plots: non-constant variance!

par(mfrow = c(1, 1))
hist(LM$residuals, breaks = 10, xlim = c(-1,1))
qqPlot(LM$residuals, line = "quartiles") # normal
shapiro.test(LM$residuals) # indicates normality of residuals

ncvTest(LM) # homoscedasticity test: H0 of constant variance is not rejected. 
# Supported by flat-ish slope in Scale-location plot (plot(testLM))

influenceIndexPlot(LM) # outliers test, nothing concerning

boxCox(LM) # recommended to stay as is
```

* Even spread of residuals around 0
* t-values are far from 1 and both are significant
* Residual standard error is very high compared to the estimate
* Only approx. 17% of the variance of fst.dist can de explained by geo.dist!

# 8. Examining values below the regression line

```{r summary-stats}
mean(pwd.df$fst.dist)
median(pwd.df$fst.dist)
range(pwd.df$fst.dist)
# without new guinea and norm. isl.:
mean(pwd.df2$fst.dist)
median(pwd.df2$fst.dist)
range(pwd.df2$fst.dist)
```

```{r values-under-line}
fv <- as.vector(LM$fitted.values)
under.fv <- cbind(pwd.df2, fv)

under.fv <- under.fv[under.fv$fst.dist < under.fv$fv,] # keeping only fst.dist 
# values less than fitted values

plot(fst.dist ~ geo.dist, data = under.fv)
# should be all values up to the regression line
```

```{r under-line-table}
under.fv <- under.fv[order(under.fv$geo.dist, decreasing = FALSE),] # sorting by geo.dist
under.fv <- under.fv[order(under.fv$fst.dist, decreasing = FALSE),] # sorting by fst.dist
under.fv <- under.fv[order(under.fv$islands.combo, decreasing = FALSE),] # sorting alphabetically
# kable(under.fv)

rownames(under.fv) <- seq(1:nrow(under.fv))
under.fv.copy <- tidyr::separate(under.fv, sep = ":", col = islands.combo, into = c("isl1", "isl2"))
```

```{r mapping-prep-1}
# this chunk is dedicated to creating a df like this for a plot with connecting points:
# island  lat long  group
# aotea   123 321   aotea:borneo
# borneo  345 543   aotea:borneo
# aotea   123 321   aotea:kaikura
# kaikura 567 765   aotea:kaikura

under.fv.copy$islands.combo <- under.fv$islands.combo # adding island.combo to define where the line on the plot needs to go between (the "group")
under.fv.copy <- rbind(under.fv.copy, under.fv.copy) # making 2 of each island combo
nrow(under.fv.copy)
under.fv.copy[153:306,1] <- under.fv.copy[153:306,2] # doing my own merge of the first 2 columns
under.fv.copy <- under.fv.copy[,-2] # now removing the 2nd column now that I fixed the 1st column
```

```{r mapping-prep-2}
head(longlat) # checking correct columns are used
dim(longlat)

longlat[longlat$geo_long < 0,] # only longitude values up to 180 are plotting, rest 
# is blank. Can try fixing this by adding 360 to all the negative longitude points
longlat[6,3] <- longlat[6,3] + 360
longlat[7,3] <- longlat[7,3] + 360
longlat[9,3] <- longlat[9,3] + 360
longlat[11,3] <- longlat[11,3] + 360
longlat[14,3] <- longlat[14,3] + 360
longlat[20,3] <- longlat[20,3] + 360
longlat[21,3] <- longlat[21,3] + 360
longlat[24,3] <- longlat[24,3] + 360

# adding lat and long to df
under.fv.copy <- merge(under.fv.copy, longlat, by.x = "isl1", by.y = "island.1") 
# sorting by fst
under.fv.copy <- under.fv.copy[order(under.fv.copy$fst.dist, decreasing = TRUE),] 

rownames(under.fv.copy) <- seq(1:nrow(under.fv.copy))
# z$islands.combo
under.fv.copy <- dplyr::mutate(under.fv.copy, group.id = match(islands.combo, unique(islands.combo)))
```

```{r mapping}
# library(ggrepel)
range(under.fv.copy$fst.dist) # checking what the midpoint would be for coloured legend

ggplot(under.fv.copy, aes(x = geo_long, y = geo_lat, group = group.id, colour = fst.dist)) +
  borders("world2", colour = NA, fill = "grey85")  +
  geom_path(size = 0.3) +
  scale_color_gradient2(
    low = "blue",
    mid = "green",
    high = "red",
    midpoint = 0.28644,
    name = "FST value"
  ) +
  geom_point(color = "black", size = 1) +
  scale_x_continuous(limits = c(100, 240)) +
  scale_y_continuous(limits = c(-50, 35)) +
  theme(panel.background = element_rect(fill = "white", colour = "white"), 
        axis.text = element_blank(), 
        axis.title = element_blank(),
        axis.ticks = element_blank()) +
  # geom_text_repel(size = 3, point.size = 1, min.segment.length = 0.25) +
  ggtitle(
    "Map of the Island Pairs with FST Values lower than expected given Geographic Distance")
```

```{r mapping-smaller-df}
# map of lines that are more than 0.15 lower than the regression line
x <- under.fv[under.fv$fst.dist < under.fv$fv - 0.15,] # making df with lowest fst values
foo <- x$islands.combo
x <- tidyr::separate(x, sep = ":", col = islands.combo, into = c("isl1", "isl2"))
x$islands.combo <- foo # adding island.combo to define where the line on the plot needs to go between (the "group")
rm(foo)
x <- rbind(x, x) # making 2 of each island combo
rownames(x) <- seq(1:nrow(x))
nrow(x)
x[43:86,1] <- x[43:86,2] # doing my own merge of the first 2 columns
x <- x[,-2] # now removing the 2nd column now that I fixed the 1st column
x <- merge(x, longlat, by.x = "isl1", by.y = "island.1") # adding lat and long to df
x <- x[order(x$fst.dist, decreasing = TRUE),] # sorting by fst
rownames(x) <- seq(1:nrow(x))
# x$islands.combo
x <- dplyr::mutate(x, group.id = match(islands.combo, unique(islands.combo))) # making 
# group id number column

range(x$fst.dist)

ggplot(x, aes(x = geo_long, y = geo_lat, group = group.id, colour = fst.dist)) +
  borders("world2", colour = NA, fill = "grey85")  +
  geom_path(size = 0.3) +
  scale_color_gradient2(
    low = "blue",
    mid = "green",
    high = "red",
    midpoint = 0.214025,
    name = "FST value"
  ) +
  geom_point(color = "black", size = 1) +
  scale_x_continuous(limits = c(100, 240)) +
  scale_y_continuous(limits = c(-50, 35)) +
  theme(panel.background = element_rect(fill = "white", colour = "white"), 
        axis.text = element_blank(), 
        axis.title = element_blank(),
        axis.ticks = element_blank()) +
  # geom_text_repel(size = 3, point.size = 1, min.segment.length = 0.25) +
  ggtitle(
    "Map of the Island Pairs with FST Values far lower than expected given Geographic Distance")
```

```{r everything-map}
# map of all connecting points
foo <- pwd.df2$islands.combo
y <- tidyr::separate(pwd.df2, sep = ":", col = islands.combo, into = c("isl1", "isl2"))
y$islands.combo <- foo # adding island.combo to define where the line on the plot 
# needs to go between (the "group")
rm(foo)
y <- rbind(y, y) # making 2 of each island combo
rownames(y) <- seq(1:nrow(y))
nrow(y)
y[298:596,1] <- y[298:596,2] # doing my own merge of the first 2 columns
y <- y[,-2] # now removing the 2nd column now that I fixed the 1st column
y <- merge(y, longlat, by.x = "isl1", by.y = "island.1") # adding lat and long to df
y <- y[order(y$fst.dist, decreasing = TRUE),] # sorting by fst
rownames(y) <- seq(1:nrow(y))
# x$islands.combo
y <- dplyr::mutate(y, group.id = match(islands.combo, unique(islands.combo))) # making 
# group id number column

range(y$fst.dist)

ggplot(y, aes(x = geo_long, y = geo_lat, group = group.id, colour = fst.dist)) +
  borders("world2", colour = NA, fill = "grey85")  +
  geom_path(size = 0.3) +
  scale_color_gradient2(
    low = "blue",
    mid = "green",
    high = "red",
    midpoint = 0.419645,
    name = "FST value"
  ) +
  geom_point(color = "black", size = 1) +
  scale_x_continuous(limits = c(100, 240)) +
  scale_y_continuous(limits = c(-50, 35)) +
  theme(panel.background = element_rect(fill = "white", colour = "white"), 
        axis.text = element_blank(), 
        axis.title = element_blank(),
        axis.ticks = element_blank()) +
  ggtitle(
    "Map of the Island Pairs with FST Values between populations")
```

# 9. Plots 

```{r correlogram, eval = FALSE}
library(corrplot)
par(mfrow = c(1, 1))
corrplot(pwd, 
         method = "color", 
         type = "lower", # which triangle
         tl.col = "black", # text colour
         order = "FPC",
         diag = FALSE,
         p.mat = pv, # links to p-value matrix
         sig.level = 0.05,
         insig = "pch", # what to do with insignificant p-values
         col.lim = c(-0.4, 1),
         col = RColorBrewer::brewer.pal(n = 10, name = "Spectral") # colour palette
         )
mtext("Correlogram of FST values between Islands", at = 10, line = -5, cex = 1.3) 
# at = horizontal, line = height, cex = size

# this plot is best seen in the viewer rather than in the knitted pdf. 
# Has also been saved as a separate file.
```

From 0 to 1: 0 implying free interbreeding, 1 means the popn.s don't interbreed
"Values for mammal populations between subspecies, or closely related species, typical values are of the order of 5% to 20%"

```{r ggplot-fst}
ggplot(data = pwd.df2, aes(x = geo.dist, y = fst.dist)) + 
  geom_point(shape = 1, colour = "black") + 
  geom_smooth(method = "lm", se = TRUE) + 
  ggtitle("Correlation between FST and Geographic Distance between Islands") +
  xlab("Geographic Distance between two islands (km)") + 
  ylab("FST value between two islands") +
  theme_light()
```

The two negative values may be due to low sample size (could be outliers). FST is usually between 0 and 1. The negative values are also identified and non-significant.

```{r ggplot-fst-2}
fv <- as.vector(LM$fitted.values)
pwd.df2 <- cbind(pwd.df2, fv)
z <- dplyr::mutate(pwd.df2, colour = ifelse(fst.dist < (fv - 0.15), "red", "black"))

ggplot(data = z, aes(x = geo.dist, y = fst.dist)) + 
  geom_point(aes(colour = colour), shape = 1) +
  scale_color_identity() +
  geom_smooth(method = "lm", se = TRUE) + 
  ggtitle("Correlation between FST and Geographic Distance between Islands") +
  xlab("Geographic Distance between two islands (km)") + 
  ylab("FST value between two islands") +
  theme_light()
```

