---
title: "Mantel Test Script"
author: "Grace Saville"
date: "16/05/2022"
output: pdf_document
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE)
opts_knit$set(root.dir = "C:/Users/airhe/OneDrive/Documents/Masters/Project 3/kiore-project")
```

```{r libraries, include = FALSE}
library(ade4)
library(geosphere)
library(ggplot2)
library(car)
library(plyr)
# getwd()
# setwd("C:/Users/airhe/OneDrive/Documents/Masters/Project 3/kiore-project")
```

# 1. Loading the data

```{r loading-data}
data <- read.csv("./data/RStudio/ratsSNPs_clean.csv")
gen.matrix <- as.matrix(
  read.delim("./data/SplitsTree/geneticdist_SplitsTree_output.txt", 
             sep = "\t", header = FALSE))
# Make sure the text file is the distances only, remove any extras e.g. 
# column vector at bottom of file

dim(data)
```

# 2. Creating geographical distance matrix

**ONLY NEED TO RUN THE FOLLOWING CODE CHUNK ONCE AND THEN SAVE**

Using the mapview funtion in the last code chunk of this script I noticed that Tahanea is plotted in Australia, and found out the longitude number is missing a negative sign, and that Reiono and Honuea have the same coordinates even though they're different islands. Fixing that here:
```{r longitude-adjustment, eval = FALSE}
x <- grep("Tahanea", data$island.1, value = FALSE) # finding rows of Tahanea
names(data) # finding column number for "geo_long"
data[x, 12]
data[x, 12] <- -144.97 # replacing the number

# the coordinates given for Reiono and Honuea point to an island in 
# French Polynesia called Moorea-Maiao
# Reino should be approx. -17.046, -149.546
# Honuea should be approx -17.009, -149.585
# Checking other islands close by: 
# Rimatu'u is -17, -149.57 (over the sea) but should be approx -17.03, -149.558

x <- grep("Reiono", data$island.1, value = FALSE)
data[x, c(11, 12)]
data[x, 11] <- -17.046 # replcing geo_lat
data[x, 12] <- -149.546 # replacing geo_long

x <- grep("Honuea", data$island.1, value = FALSE)
data[x, c(11, 12)]
data[x, 11] <- -17.009 # replcing geo_lat
data[x, 12] <- -149.585 # replacing geo_long

x <- grep("Rimatuu", data$island.1, value = FALSE)
data[x, c(11, 12)]
data[x, 11] <- -17.03 # replcing geo_lat
data[x, 12] <- -149.558 # replacing geo_long

rm(x)

# write.csv(data, "./data/RStudio/ratsSNPs_clean.csv", row.names = FALSE)
```

(only need to do the above code once since it saves the edited df to file)

# 3. Setting up longitudes and latitudes

Different distance functions:

* distHaversine() assumes earth is a sphere
* distm() makes distance matrix
* distGeo() assumes earth is elliptical (ish), can choose specific model

```{r longitude-latitude-matrix}
head(names(data), n = 20) # need "geo_lat" "geo_long" 
longlat <- data[,c(1,11,12)]
head(longlat) # checking correct columns are used

longlat <- as.matrix(longlat[,c(3,2)]) # distGeo function needs a matrix 
# with 2 columns, col 1 longitude and col 2 latitude

geo.matrix <- distm(longlat, fun = distGeo) # converting to pairwise distance matrix
dim(geo.matrix) # 370 370

geo.dist <- as.dist(geo.matrix, diag = TRUE, upper = TRUE) # converting to dist object
# diag = TRUE #includes diagonal zeros
# upper = TRUE #includes upper triangle
```

# 4. Creating genetic distance matrix

```{r genetic-distance-matrix}
dim(gen.matrix) # 370 370
gen.dist <- as.dist(gen.matrix, diag = TRUE, upper = TRUE) # converting to dist object
```

# 5. Mantel test

```{r mantel-test, warning = FALSE}
set.seed(4)
r1 <- mantel.rtest(gen.dist, geo.dist, nrepet = 999)
r1
plot(r1$plot$hist, main = "Mantel test", xlim = c(-0.1, 0.1)) # plotting
# abline(v = r1$obs, col = 4, lwd = 2)

r1$obs^2 # r squared
```

* -1 suggests strong negative correlation, e.g. closer islands mean further genetically or further islands means closer genetically
* 0 suggests no correlation, e.g. genetic difference is not correlated to island distance
* 1 suggests strong positive correlation e.g. closer islands mean closer genetically 

Therefore the observed correlation of 0.4987612 suggests that there is a positive correlation between genetic distance and geographic distance (and the null hypothesis of no correlation is rejected).

# 6. Creating pairwise distances dataframe

```{r making-pwd-df}
specimens <- read.delim(
  "./data/SplitsTree/geneticdist_SplitsTree_output_taxa_only.txt", header = FALSE)
specimens <- as.vector(specimens[,2])

colnames(gen.matrix) <- specimens 
rownames(gen.matrix) <- specimens # naming the rows and columns by the order 
# given in the SplitsTree output

colnames(geo.matrix) <- data[,1]
rownames(geo.matrix) <- data[,1] # naming the rows and columns here by the 
# order of lat/long, which came from the data df

gen.matrix[lower.tri(gen.matrix)] <- NA # keeping only the upper triangle of each matrix
geo.matrix[lower.tri(geo.matrix)] <- NA

dist.summary <- data.frame(
  col = colnames(gen.matrix)[col(gen.matrix)],
  row = rownames(gen.matrix)[row(gen.matrix)],
  gen.dist = c(gen.matrix)
) # converting the genetic matrix into a df with columns describing which 
# combos result in the distance

x <- data.frame(
  col = colnames(geo.matrix)[col(geo.matrix)],
  row = rownames(geo.matrix)[row(geo.matrix)],
  geo.dist = c(geo.matrix)
) # doing the same with the geographic matrix

dist.summary <- merge(dist.summary, x, by = 1:2, all = TRUE) # merging the 2 dfs
rm(x)
```

The above df "dist.summary" is not ideal since it's not space efficient (~12MB), however it provides and easy way to link the row and column specimens that generated the distances, therefore making points on the graph label-able.

```{r tidying-pwd-df}
dist.summary <- tidyr::unite(dist.summary, specimens.combo, 1:2, sep = ":", remove = TRUE)
# combining the first 2 columns (the names of the matrices columns and rows) 
# to give a label to each pairwise distance

dist.summary <- na.omit(dist.summary) # removing NA's that were in the bottom triangle

# creating a column of island combination labels, rather than specimen combination:
islands.combo <- dist.summary$specimens.combo 
islands.combo <- gsub("[0-9]+", "", islands.combo) # removing specimen ID numbers
islands.combo <- gsub("_", "", islands.combo) # removing underscores
# unique(islands.combo)
islands.combo <- gsub("Thailand", "Mainland", islands.combo) # replacing 3 
# countries with mainland
islands.combo <- gsub("Laos", "Mainland", islands.combo)
islands.combo <- gsub("Cambodia", "Mainland", islands.combo)
islands.combo <- gsub("Halmahera", "Halmaher", islands.combo) # replacing some double ups
islands.combo <- gsub("NewBritai", "NewBrita", islands.combo)
islands.combo <- gsub("NewGuinea", "NewGuine", islands.combo)
islands.combo <- gsub("Motukawan", "Motukawa", islands.combo)
dist.summary <- cbind(dist.summary, islands.combo) # adding this new column
rm(islands.combo)

# removing self-self distances, since I'm looking at inter-island not intra-island
intra <- as.character(c(
  "Aotea:Aotea",
  "Borneo:Borneo",
  "GrtMercury:GrtMercury",
  "Halmaher:Halmaher",
  "Hatutaa:Hatutaa",
  "Honuea:Honuea",
  "Kaikura:Kaikura",
  "Kamaka:Kamaka",
  "Kayangel:Kayangel",
  "LateIs:LateIs",
  "Luzon:Luzon",
  "Mainland:Mainland",
  "Malenge:Malenge",
  "Mohotani:Mohotani",
  "Motukawa:Motukawa",
  "NewBrita:NewBrita",
  "NewGuine:NewGuine",
  "Normanby:Normanby",
  "Rakiura:Rakiura",
  "Reiono:Reiono",
  "Rimatuu:Rimatuu",
  "Slipper:Slipper",
  "Southland:Southland",
  "Sulawesi:Sulawesi",
  "Tahanea:Tahanea",
  "WakeIs:WakeIs"
))
dist.summary[which(dist.summary$islands.combo %in% intra),4] <- NA # removing 
# same-same island combos
dist.summary[dist.summary == 0] <- NA # turning zeros to NAs
dist.summary <- na.omit(dist.summary) # removing rows with NAs
rm(intra, specimens)

dist.summary$geo.dist <- dist.summary$geo.dist/1000 # going from metres to km
```

Rows which have either or both gen and geo dist at zero (although there should never be a gen.dist at zero) will sit on the axis and potentially pull on the regression line. Geo distances at 0 will always be specimens on the same island, which I think it best to remove since they don't contribute to whether there is a correlation between islands distance (since they're actually the same island).

```{r saving-pwd-df, eval = FALSE}
# getwd()
write.csv(dist.summary, "./data/RStudio/gen_geo_distance_matrices_df.csv", row.names = FALSE)
```

# 7. Linear Modelling

```{r reloading-data, eval = FALSE}
dist.summary <- read.csv("./data/RStudio/gen_geo_distance_matrices_df.csv")
```

## 7a. Test Model

```{r test-model-1-building}
testLM <- lm(gen.dist ~ geo.dist, data = dist.summary) # model
summary(testLM) # model results

plot(gen.dist ~ geo.dist, data = dist.summary)
abline(coef = coef(testLM), col = 4, lwd = 2)
```

```{r test-model-1-diagnostics}
par(mfrow = c(2, 2)) # changes the number of plots visible at once
plot(testLM) # # diagnostic plots, not normal, distribution may be skewed 
# towards the left side ("Right skew" or "positive skew")

par(mfrow = c(1, 1))

qqPlot(testLM$residuals, line = "quartiles") # non-normal dist
ks.test(testLM$residuals, 'pnorm') # non-normal
hist(testLM$residuals, breaks = 50) # clear positive skew

ncvTest(testLM) # homoscedasticity test: H0 of constant variance is rejected

influenceIndexPlot(testLM) # outliers, hard to judge since there seem to be many, 
# or none which are an issue
outlierTest(testLM) # lists 10 different points which could be an issue 
# (all over 3 rstudent and signif.)

boxCox(testLM) # recommended log transformation
```

I decided to do a brief check of observed correlation to compare with the Mantel test results, using the above dataframe;
```{r additonal-correlation-test}
cor.test(dist.summary$gen.dist, dist.summary$geo.dist, method = "pearson") # 
# should be taken with a grain of salt since the residual variance is not constant
```

## 7b. Adjusted Model

I attempted to transform the data (e.g. sqrt, log10) to achieve normal residual distribution however it was not successful so I will build a GLM instead.

```{r adjusted-model-1-building}
GLM <- glm(gen.dist ~ geo.dist, data = dist.summary, family = Gamma(link = "identity")) 
# Gamma distrubution family used for right-skewed data. 
# I tried the link functions "log" and "inverse" first but the resulting coefficients 
# were very peculiar (e.g. intercepts at -2 or 7)
summary(GLM) # model results

par(mfrow = c(1, 1))
plot(gen.dist ~ geo.dist, data = dist.summary)
abline(coef = coef(GLM), col = 4, lwd = 2)
```

```{r adjusted-model-1-diagnostics}
par(mfrow = c(2, 2)) # changing the number of plots visible at once 
plot(GLM) # diagnostic plots

residualPlots(GLM)

par(mfrow = c(1, 1))
qqPlot(GLM$residuals, line = "quartiles") # 
ks.test(GLM$residuals, 'pnorm') # 
hist(GLM$residuals, breaks = 50)
```

```{r variance-model-explains}
with(summary(GLM), 1 - deviance/null.deviance) # R^2 value of 0.2584382. 
# This is not high therefore there may be additional variables to add to 
# the model to explain the variance.
```

# 8. Looking at the genetically closest island pairs (based on specimens)
```{r genetically-closest, eval = FALSE}
# top 10 genetically closest islands
# sorting by geo dist, increasing in size
dist.summary <- dist.summary[order(dist.summary$geo.dist, decreasing = FALSE),] 

# sorting by gen dist, 
dist.summary <- dist.summary[order(dist.summary$gen.dist, decreasing = FALSE),] 


x <- dist.summary[!duplicated(dist.summary$islands.combo),]

x <- x[order(x$gen.dist, decreasing = FALSE),]
kable(head(x, n = 20))
```

# 9. Taking a closer look at islands with high distance/close genetics

```{r values-under-line}
# reloading the df so that the order is reset to original:
dist.summary <- read.csv("./data/RStudio/gen_geo_distance_matrices_df.csv")
# making sure correct fitted values are ready:
GLM <- glm(gen.dist ~ geo.dist, data = dist.summary, family = Gamma(link = "identity"))

fv <- as.vector(GLM$fitted.values)
x <- cbind(dist.summary, fv)

x <- x[x$gen.dist < x$fv,] # keeping only gen.dist values less than fitted values

plot(gen.dist ~ geo.dist, data = x)
# should be all values up to the regression line
```

```{r}
all.combos <- count(dist.summary$islands.combo) # if doesn't work try $ instead of comma
under.fv.combos <- count(x$islands.combo)
combo.totals <- merge(all.combos, under.fv.combos, by = "x", all = TRUE)
colnames(combo.totals) <- c("combo", "all", "below.line")
rm(all.combos, under.fv.combos, x)
combo.totals$perc.under <- round(((combo.totals$below.line / combo.totals$all) * 100), 3)
combo.totals[is.na(combo.totals)] <- 0

# perc.under is the percentage of the specimens with genetic distances (between 
# the two stated islands) which are less than the average genetic distance/point 
# on the regression line at that geographic distance.

kable(combo.totals[combo.totals$perc.under > 99,])

# x <- combo.totals$combo[combo.totals$perc.under == 100]
# x <- as.data.frame(x)
# x <- tidyr::separate(x, sep = ":", col = x, into = c("isl1", "isl2"))
```

# 11. Plots

```{r plots}
ggplot(data = dist.summary, aes(x = geo.dist, y = gen.dist)) + 
  geom_point(shape = 1, colour = "grey") + 
  geom_smooth(method = "lm") + 
  ggtitle("Correlation between Geographic and Genetic Distance between Islands") +
  xlab("Geographic Distance between two islands (km)") + 
  ylab("Genetic Distance between two islands") +
  theme_light()
```

# 11. Plotting the coordinates on a map

```{r include = FALSE}
library(maps)
```

```{r mapping}
head(names(data), n = 20) # need "geo_lat" "geo_long" 
longlat <- data[,c(8,11,12)]
longlat <- longlat[!duplicated(longlat$island.1),] # keeping only 1 coordinate for each island
head(longlat) # checking correct columns are used
dim(longlat)
rownames(longlat) <- 1:26

longlat$island.1
longlat[14,1] <- "Rakiura"
longlat[18,1] <- "Rimatuu"
longlat[19,1] <- "Aotea"

longlat[longlat$geo_long < 0,] # only longitude values up to 180 are plotting, rest 
# is blank. Can try fixing this by adding 360 to all the negative longitude points
longlat[8,3] <- longlat[8,3] + 360
longlat[9,3] <- longlat[9,3] + 360
longlat[11,3] <- longlat[11,3] + 360
longlat[13,3] <- longlat[13,3] + 360
longlat[15,3] <- longlat[15,3] + 360
longlat[16,3] <- longlat[16,3] + 360
longlat[18,3] <- longlat[18,3] + 360
longlat[22,3] <- longlat[22,3] + 360


map("world2", # "world" over Africa, "world2" over Pacific
    xlim = c(90,240), # x axis limits
    ylim = c(-70,70),
    fill = TRUE, 
    col = "grey", # fill colour
    # bg = "azure2", # background colour
    border = NA, # country borders
    wrap = TRUE # tuck map into axes
    ) 
map.axes() # adds axes
points(longlat$geo_long, longlat$geo_lat, 
       col = "red", cex = 1, pch = 20)
```

```{r include = FALSE}
library(ggrepel)
```


```{r ggplot-map}
# trying with ggplot
ggplot(longlat, aes(x = geo_long, y = geo_lat, label = island.1)) +   
  borders("world2", colour = NA, fill = "wheat1")  +
  geom_point(color = "red", size = 2) +
  scale_x_continuous(limits = c(100, 240)) +
  scale_y_continuous(limits = c(-50, 35)) +
  theme(panel.background = element_rect(fill = "azure1", colour = "azure1"), 
        axis.text = element_blank(), 
        axis.title = element_blank(),
        axis.ticks = element_blank()) +
  geom_text_repel(size = 3, point.size = 1, min.segment.length = 0.25) +
  ggtitle("Map of Islands Sampled for Polynesian Rats")
```







