---
title: "Mantel Test Script"
author: "Grace Saville"
date: "01/02/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

# Preamble

```{r libraries}
library(ade4)
library(geosphere)
getwd()
setwd("C:/Users/airhe/OneDrive/Documents/Masters/Project 3/kiore-project")
```

# Loading the data

```{r loading data}
data <- read.csv("./data/ratsSNPs_clean.csv")
gen.dist <- read.delim("./data/geneticdist_SplitsTree_output.txt", sep = "\t", header = FALSE)
# Make sure the text file is the distances only, remove any extras e.g. column vector at bottom of file
```

# Creating geographical distance matrix

Different distance functions:

* distHaversine() assumes earth is a sphere
* distm() makes distance matrix
* distGeo() assumes earth is elliptical (ish), can choose specific model

```{r longitude and latitude matrix}
names(data) # need "geo_lat" "geo_long" 
longlat <- data[,c(1,11,12)]
head(longlat) # checking correct columns are used

longlat <- as.matrix(longlat[,c(3,2)]) # distGeo function needs a matrix with 2 columns, col 1 longitude and col 2 latitude

geo.dist <- distm(longlat, fun = distGeo) # converting to pairwise distance matrix
dim(geo.dist) # 370 370

geo.dist <- as.dist(geo.dist, diag = TRUE, upper = TRUE) # converting to dist object
# diag = TRUE #includes diagonal zeros
# upper = TRUE #includes upper triangle
```

# Creating genetic distance matrix

```{r genetic distance matrix}
gen.dist <- as.matrix(gen.dist)
dim(gen.dist) # 370 370
gen.dist <- as.dist(gen.dist, diag = TRUE, upper = TRUE) # converting to dist object
```

# Mantel test

```{r mantel test}
mantel.rtest(gen.dist, geo.dist, nrepet = 999)
```

**Results:**  
Monte-Carlo test  
Call: mantelnoneuclid(m1 = m1, m2 = m2, nrepet = nrepet)  

Observation: 0.4345602   

Based on 999 replicates  
Simulated p-value: 0.001   
Alternative hypothesis: greater   

Std.Obs: 2.329675e+01   
Expectation: 7.916798e-04     
Variance: 3.466772e-04  

* -1 suggests strong negative correlation, e.g. closer islands mean further genetically or further islands means closer genetically
* 0 suggests no correlation, e.g. genetic difference is not correlated to island distance
* 1 suggests strong positive correlation e.g. closer islands mean closer genetically 

Therefore the observed correlation of 0.4345602 suggests that there is a positive correlation between genetic distance and geographic distance (and the null hypothesis of no correlation is rejected).


**Results from uncleaned dataset:**  
Monte-Carlo test  
Call: mantelnoneuclid(m1 = m1, m2 = m2, nrepet = nrepet)  

Observation: 0.2807331

Based on 99 replicates  
Simulated p-value: 0.01  
Alternative hypothesis: greater   

Std.Obs: 26.5367570885  
Expectation: -0.0002961658  
Variance: 0.0001121521  

# Plots

```{r plots}
x <- data.frame(as.vector(gen.dist), as.vector(geo.dist))
colnames(x) <- c("Genetic_Distance", "Geographic_Distance")

library(ggplot2)
ggplot(data = x, aes(x = Geographic_Distance, y = Genetic_Distance)) + geom_point(shape = 1, colour = "grey") + geom_smooth(method = "lm") + geom_quantile(quantiles = c(0.05,0.95)) 
# linear smooth regression line and 5/95 quantile lines added
```







