---
title: "Creating file for NeighborNet Treebuilding"
author: "Grace Saville"
date: "01/10/2021"
output: pdf_document
---
# 1. Preamble 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, tidy.opts = list(width.cutoff = 70), tidy = TRUE)
```

```{r libraries}
library(tidyr)
library(ape)
library(phylotools)
getwd()
setwd("C:/Users/airhe/OneDrive/Documents/Masters/Project 3/kiore-project")
```

# 2. Loading the data

```{r data loading}
data <- read.csv("./data/RStudio/ratsSNPs_clean.csv")
dim(data) # 370 rows 298 columns
str(data)
```

# 3. Testing the dat2phylip() function
```{r testing}
# x <- data.frame(a = c("asdfghjkl1", "asdf2ghjkl", "asdf3ghjkl", "asdfghjkl4"), b = c("CTAGTGACCCGTAG","TGACCCGTTAGAAC", "TGACTCTTTAGAAC", "TTTCACGTTGAGAC")) # dummy df

# dat2phylip(x, outfile = "test.phy") # testing this function that saves phylip files
```

After some experimenting, for Phylip files...

* spaces in the base strings aren't necessary
* 10 character length names are needed
* dashes "-" are allowed
* no differentiation between upper and lower case
* only letters from bases or amino acids allowed (?, e.g acdefghiklmnprstvwyz)


# 4. Replacing the SNP symbols with IUPAC Ambiguity code symbols

IUPAC Ambiguity code:  

|Symbol|SNP bases|
|:----:|:----:|
|A|AA|
|T|TT|
|C|CC|
|G|GG|
|R (purine)|AG|
|Y (pyrimidine)|CT|
|W (weak)|AT|
|S (strong)|CG|
|M (amino)|AC|
|K (keto)|GT|

```{r replacing with IUPAC code}
copy <- data # making a copy
table(unlist(copy[,-c(1:16)])) # ?, A:A, A:C, A:G, A:T, C:C, C:G, C:T, G:G, T:G, T:T

copy[copy == "A:A"] <- "A"
copy[copy == "T:T"] <- "T"
copy[copy == "C:C"] <- "C"
copy[copy == "G:G"] <- "G"
copy[copy == "A:G"] <- "R"
copy[copy == "C:T"] <- "Y"
copy[copy == "A:T"] <- "W"
copy[copy == "C:G"] <- "S"
copy[copy == "A:C"] <- "M"
copy[copy == "T:G"] <- "K"

table(unlist(copy[,-c(1:16)])) # ?, A, C, G, K, M, R, S, T, W, Y 
```

# 5. Adjusting the data frame for Phylip output function

```{r df adjustment}
names(copy)
copy <- copy[,-c(2:16)] # removing the rows with information other than the species key and codes
copy <- unite(copy, bases, -1, sep = "", remove = TRUE) # merging all the base columns into one
```

```{r phylip file creation}
dat2phylip(copy, outfile = "./data/SplitsTree/ratsSNPs_SplitsTree_input.phy") 
# can be opened by splitstree when configured as proteins/amino acids
```

